\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-11 pldi/draft.tex
Pass over section 4 done

@@

Often these commands do not operate in an online fashion, but need to block until the end of a stream.
A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
Such constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $reduce$ function.

@@

In addition to the commands in \sta that can be parallelized in a
straightforward way, there also exists a subclass of commands in \pur
that can be parallelized using divide-and-conquer techniques (as
mentioned in \cref{cmd}). \kk{Maybe show an example of map-reduce for
  sort or wc. Either here or in \cref{cmd}.} In order to support the
parallelization of arbitrary commands in this category, \sys allows
command developers to provide custom map and reduce steps for them,
where the map is in \sta and the reduce in \pur, and their composition
produces the same output with the initial command. \sys can then
replace occurrences of these commands with their map and reduce parts,
enabling their parallelization. \todo{We have defined these primitives
  for most of the parallelizable pure commands in coreutils, providing
  a foundation for this effort.}

@@

\kk{Find a better name.} \todo{One-input-output} commands \todo{are
  special commands} that consume their inputs one at a time, with the
possible exception of static input files, and produce one output
file. These commands can be represented as functions of type $f :
D\kstar \times [D\kstar] \to D\kstar$, where the first argument
represents the concatenation of the inputs that are consumed one at a
time and the second argument represents the static inputs. The
quintessential example of \todo{one-input-output} commands is
\ttt{cat}, that consumes its inputs one at a time, producing their
concatenation as output.
%% A \todo{large subset} of shell commands fall in this category

@@

Having a dataflow graph representation allows us to define
parallelization optimizations as graph transformations. Before
describing the different types of optimizations, we formalize the
intuition about stateless and pure commands that was described
earlier~\sx{cmd}.
%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }

\kk{Maybe
  have an argument here that we show in the evaluation that handling
  these commands is adequate for a big variety of data processing
  scripts (and most stateless and pure commands), from several
  microbenchmarks to some large-realistic macrobenchmarks.}.

As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr} operate
independently on individual elements of the stream (characters, lines,
or files) without maintaining any state. In this work, we consider
lines as the data quantum, so commands that are file-stateless are not
considered stateless. Formally, a command $f$ is stateless if it is
``linear'' with respect to concatenation, \ie satisfying the following
equation:

@@

\noindent
This means that one gets the same output by applying a command $f$ to a
concatenation of two inputs $x, x'$ as by concatenating the result
produced by the application of $f$ to these inputs
separately.

Similarly, it was mentioned that some pure commands (such
as \ttt{sort} and \ttt{wc}) can be parallelized using
divide-and-conquer techniques. More formally, these pure commands $f$
can be implemented as a combination of functions map $m$ and reduce $r$, satisfying
the following equation:

\[
\forall x, x', s, f(x.x', s) = r(m(x, s).m(x', s), s)
\]

\noindent
Similarly to the earlier equation, this means that we get the same output by
applying $f$ to a concatenation of two inputs $x, x'$, or by applying
the reduce function $r$ to the concatenation of the outputs produced
by applying map $m$ to each of these inputs.

\paragraph{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-reduce pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

\begin{figure}[t]
\centering
\includegraphics[width=0.60\columnwidth]{\detokenize{./figs/dish_g1.pdf}}
\caption{
  \textbf{Stateless parallelization transformation.}
  The \texttt{cat} node is commuted with the stateless node
  to utilize available data parallelism.
  \TODO{Correct this diagram to show the cat node before
    the node and then the cat node after the stateless
    command after the optimization. Cat node should be
    different than a circle. Maybe like an ``and'' node
    in binary circuits?}
}
\vspace{-10pt}
\label{fig:parallelization-transformation}
\end{figure}

\kk{Remember to mention the assumptions that need to hold for the
  graph transformations to be valid in the Command categories
  section. Commands must be deterministic, they must not do any other
  side effect (such as writing to other files, sending signals,
  etc). However, these assumptions must already be checked when the
  developer designates the categories. Commands are assumed to always
  terminate gracefully. \sys changes the termination behaviour of
  commands as a sequential command that would exit prematurely might
  not exit completely if parallelized.}

@@

\kk{I think that we should probably rewrite this paragraph with the
  ``new'' split optimization (and maybe any others that we develop
  until then) here.}

In addition to node splits, we can also perform edge splits in order
to increase the number of incoming edges for some nodes to further expose
possible data parallelism. In cases where the edge refers to a real file, the
edge split represents splitting the real file in chunks. In
cases where the edge connects two nodes, the split effectively refers to several edges
with a bounded number of lines. Note that the final output edge of a
valid node in the graph must always be unbounded if it does not refer
to a file.

>>>>

\kk{Do we want to get in more detail about requirements? Commands must
  be deterministic, they must not do any other side effect (sending
  signals, etc). These assumptions must be checked when the developer
  designates the categories. Commands are assumed to always terminate
  gracefully. \sys changes the termination behaviour of commands as a
  sequential command that would exit prematurely might not exit
  completely if parallelized.}

@@

Often these commands do not operate in an online fashion, but need to block until the end of a stream.
A typical example of this is \ttt{sort}, which cannot start emitting results before the last input element has been consumed.
Such constraints affect task parallelism, but not data parallelism:
  \ttt{sort} can be parallelized significantly using divide-and-conquer techniques---that is, by encoding it as a combination of a $map$ and a $fold$ function.

@@

In addition to the commands in \sta that can be parallelized in a
straightforward way, there also exists a subclass of commands in \pur
that can be parallelized using divide-and-conquer techniques (as
mentioned in \cref{cmd}). \kk{Maybe show an example of map-fold for
  sort or wc. Either here or in \cref{cmd}.} In order to support the
parallelization of arbitrary commands in this category, \sys allows
command developers to provide custom map and fold steps for them,
where the map is in \sta and the fold in \pur, and their composition
produces the same output with the initial command. \sys can then
replace occurrences of these commands with their map and fold parts,
enabling their parallelization. \todo{We have defined these primitives
  for most of the parallelizable pure commands in coreutils, providing
  a foundation for this effort.}

@@

\kk{Find a better name.} \todo{One-input-output} commands \todo{are
  special commands} that consume their inputs one at a time, with the
possible exception of static input files, and produce one output
file. These commands can be represented as functions of type $f :
D\kstar \times [D\kstar] \to D\kstar$, where the first argument
represents the concatenation of the inputs that are consumed one at a
time and the second argument represents the static inputs. The
quintessential example of \todo{one-input-output} commands is
\ttt{cat}, that consumes its inputs one at a time, producing their
concatenation as output. We focus on these commands because a large
subset of \sta and \pur commands fall in this category.

@@

We have defined a set of semantics-preserving graph transformations
that act as parallelization-exposing optimizations. Both the domain
and range of these transformations is a graph in our dataflow model
and therefore they can be composed arbitrarily and in any order.
%% Having a dataflow graph representation allows us to define
%% parallelization optimizations as graph transformations.
Before describing the different types of optimizations, we formalize
the intuition about stateless and pure commands that was described
earlier~\sx{cmd}.

%% We focus on \todo{one-input-output} commands (as defined in
%% \cref{graph-components}) that constitute a large subset of shell
%% commands.

%% \nv{TODOs
%%   (i) highlight earlier the antithesis between the formalisms in this section and the informalities of the previous (me)
%% }

%% \kk{Maybe
%%   have an argument here that we show in the evaluation that handling
%%   these commands is adequate for a big variety of data processing
%%   scripts (and most stateless and pure commands), from several
%%   microbenchmarks to some large-realistic macrobenchmarks.}.

\paragraph{Stateless and Parallelizable Pure Commands}

As mentioned in \Cref{cmd}, stateless commands such as \ttt{tr}
operate independently on individual elements of the stream
(characters, lines, or files) without maintaining any state. In this
work, we consider lines as the data quantum, so only commands that are
stateless regarding lines (or any finer granularity such as
characters) are considered stateless. Formally, a
\todo{one-input-output} command $f$ is stateless if it is ``linear''
with respect to concatenation, \ie satisfying the following equation:

@@

\noindent
This means that applying the command $f$ to a concatenation of two
inputs $x, x'$ produces the same output as applying $f$ to each input
$x, x'$ separately, and concatenating the outputs.

Similarly, some pure commands (such as \ttt{sort} and \ttt{wc}) can be
parallelized using divide-and-conquer techniques. More formally, these
pure commands $f$ can be implemented as a combination of functions map
$m$ and an associative fold $f'$, satisfying the following equation:

\[
\forall x, x', s, f(x.x', s) = f'(m(x, s).m(x', s), s)
\]

\noindent
This means that we get the same output by applying $f$ to a
concatenation of two inputs $x, x'$, or by applying the fold
function $f'$ to the concatenation of the outputs produced by applying
map $m$ to each of these inputs.

\paragraph{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{\detokenize{./figs/transformation-sketch.png}}
\caption{
  \textbf{Stateless parallelization transformation.}
  The \texttt{cat} node is commuted with the stateless node
  to utilize available data parallelism.
  \TODO{Correct this diagram to show the cat node before
    the node and then the cat node after the stateless
    command after the optimization. Cat node should be
    different than a circle. Maybe like an ``and'' node
    in binary circuits?}
}
\vspace{-10pt}
\label{fig:parallelization-transformation}
\end{figure}

%% \kk{Remember to mention the assumptions that need to hold for the
%%   graph transformations to be valid in the Command categories
%%   section. Commands must be deterministic, they must not do any other
%%   side effect (such as writing to other files, sending signals,
%%   etc). However, these assumptions must already be checked when the
%%   developer designates the categories. Commands are assumed to always
%%   terminate gracefully. \sys changes the termination behaviour of
%%   commands as a sequential command that would exit prematurely might
%%   not exit completely if parallelized.}

@@

\kk{Maybe say something about the rest of the transformations: (i)
  changing a command with multiple inputs to a cat followed by the
  command, (ii) Adding an identity node in the place of two edges, and
  (iii) adding a split after a command.}

%% \kk{I think that we should probably rewrite this paragraph with the
%%   ``new'' split optimization (and maybe any others that we develop
%%   until then) here.}

%% In addition to node splits, we can also perform edge splits in order
%% to increase the number of incoming edges for some nodes to further expose
%% possible data parallelism. In cases where the edge refers to a real file, the
%% edge split represents splitting the real file in chunks. In
%% cases where the edge connects two nodes, the split effectively refers to several edges
%% with a bounded number of lines. Note that the final output edge of a
%% valid node in the graph must always be unbounded if it does not refer
%% to a file.
